From physical principles to ???

### 2016-03-25 2015CompleteDerivation/completeDerivation.tex
Reviewed labeling states section

To identify and name states we will use a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen item is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variables} and \emph{state vector} a set of state variables that can identify one and only one state. We call a \emph{possibility} a possible value that can be taken by a state variable. We call a state variable \emph{discrete} or \emph{continuous} if the possibilities are integer or real numbers respectively.

We purposely use the term state variables instead of coordinates (even though that's what they are mathematically) as it will create confusion with space-time coordinates. We also avoid the term observable or measurable, as not all state variables may be directly physically meaningful (e.g. conjugate momentum in a gauge theory). The only requirements for state variables is that they identify states. This means the possible values for state variables are physically distinguishable and are defined at equal time (since states are defined and mapped at a particular time). \footnote{This is the main reason that a quantity like velocity is not a suitable state variable, as it is  defined over an interval of time. Therefore velocity is always physically meaningful but is not a state variable in general, while conjugate momentum is always a state variable but is physically meaningful in general. As we'll see later, only once a link is established between them they can both be physically meaningful state variables.}

\begin{defn}\label{state_variable}
	A \emph{state variable} is a homeomphism $q : U \rightarrow \mathbbm{L}$ where $U \subseteq \mathcal{S}$ and $\mathbbm{L}$ is the topological space for the possible values. If $\mathbbm{L}\cong \mathbbm{Z}$ the variable is said \emph{discrete}. If $\mathbbm{L}\cong \mathbbm{R}$ the variable is said \emph{continuous}.
	%We also have $\mathcal{s}_1 \neq \mathcal{s}_2 \Leftrightarrow \exists q : U \rightarrow \mathbbm{L} | q(\mathcal{s}_1)\neq q(\mathcal{s}_2) \forall \mathcal{s}_1, \mathcal{s}_2 \in U$.
\end{defn}

\begin{justification}
	We claim $\mathbbm{L}$ is a topological space. The collection of all possible values form a set. Such set is physically distinguishable. The physical outcomes induce a Hausdorff and second countable topology, as for \ref{statedef}.

	We claim $q$ is a homeomorphism. Elements of $\mathcal{S}$ and $\mathbbm{L}$ are both physically distinguishable therefore a map between them is well defined. Let $V \subseteq U$ the set of states associated with a physical outcome. $V$ is an open set in the state topology. Given the map $q$ an outcome on states can be used to distinguish possibilities of the state variables. The same outcome is associated with $g(V)$. $g(V)$ is an open set in the topology of possible values.

	%Let $\mathcal{s}_1, \mathcal{s}_2 \in \mathcal{S}$. Suppose $\mathcal{s}_1 \neq \mathcal{s}_2$. $\mathcal{s}_1$ and $\mathcal{s}_2$ are physically distinguishable. There exists a physical process with different outcomes. Let $q : \mathcal{S} \rightarrow \mathbbm{I}$ be the state variable associated to that process. Then $q(\mathcal{s}_1)\neq q(\mathcal{s}_2)$. Now suppose $\nexists q : \mathcal{S} \rightarrow \mathbbm{I} | q(\mathcal{s}_1) \neq q(\mathcal{s}_2)$. Then no process can distinguish between the two, they are physically indistinguishable therefore $\mathcal{s}_1 = \mathcal{s}_2$.
\end{justification}

When assembling multiple state variables into a state vector, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen item, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have two values. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=s$ is available; for $n=2$ we can choose $l=s$ or $l=p$. The choices are not independent. That is: two or more variables are independent if there always exists a state for any possible combination, if the total number of states is the cartesian product of the possibilities of each variable.

@@

\begin{justification}
	TODO What is there to justify?
\end{justification}

@@

The discrete variable do not present such problems, as any number of them can be flattened out in a single one (i.e. $\mathbbm{Z}^n$ is homeomorphic to $\mathbbm{Z}$). They will determine the number of connected components of the manifold. Once we introduce a continuous parameter for time evolution, though, these also become irrelevant as continuous time evolution requires continuous trajectories that cannot move states across disconnected components. This justifies the special interest in path connected manifolds, as these where trajectories for deterministic and reversible continuous evolution live.

@@

\begin{proof}
	We claim $\mathcal{S}'$ is a manifold. Let $\mathcal{s} \in \mathcal{S}$. Let $n$ be the number of independent continuous variables needed to identify $\mathcal{s}$. There exists a neighborhood $U$ around $\mathcal{s}$ where we have $(q_1,...,q_n):U\rightarrow \mathbbm{R}^n$. This map is a bijection as $s$ is identified by those variables. $\mathcal{S}$ is homeomorphic to $\mathbbm{R}^n$ around  $\mathcal{s}$. Let $\mathcal{S}'$ be the set of all states potentially reachable by deterministic and reversible evolution from $\mathcal{s}$. Deterministic and reversible evolution is a homeomorphism between initial and final states. $\forall \hat{\mathcal{s}} \in \mathcal{S}'$ the evolution of $U$ will contain $\mathcal{s}$ and will be isomorphic to $\mathbbm{R}^n$. $\mathcal{S}'$ equipped with the subspace topology is a Hausdorff, second countable topological space everywhere homeomorphic to $\mathbbm{R}^n$. $\mathcal{S}'$ is a manifold of dimension $n$.
	TODO: proof path-connectedness
\end{proof}

>>>>

To identify and name states we will use a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen item is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variables}. We call a \emph{possibility} a possible value that can be taken by a state variable. We call a state variable \emph{discrete} or \emph{continuous} if the possibilities are integer or real numbers respectively. From now on, we are going to consider state spaces whose states can be identified by a finite set of discrete and continuous state variables (i.e. the state space is locally isomorphic to $\bigcup\limits_{1 \leq i \leq n} \mathbbm{R}^{n_i}$).\footnote{TODO: Is there a topological space that is Hausdorff and second countable that is not like that?}

We purposely use the term state variables instead of coordinates (even though that's what they are mathematically) as it will create confusion with space-time coordinates. We also avoid the term observable or measurable, as not all state variables may be directly physically tangible (e.g. conjugate momentum in a gauge theory). The only requirements for state variables is that they identify states. This means the possible values for state variables are physically distinguishable and are defined at equal time (since states are defined and mapped at a particular time). \footnote{This is the main reason that a quantity like velocity is not a suitable state variable, as it is  defined over an interval of time. Therefore velocity is always physically well defined but is not a state variable in general, while conjugate momentum is always a state variable but is not physically well defined by itself (it requires the vector potential to be specified as well). As we'll see later, only once a link is established between them they can both be physically well defined state variables.}

\begin{defn}\label{state_variable}
	A \emph{state variable} is a continuous map $q : U \rightarrow \mathbbm{L}$ where $U \subseteq \mathcal{S}$ and $\mathbbm{L}$ is the topological space for the possible values. If $\mathbbm{L}\cong \mathbbm{Z}$ the variable is said \emph{discrete}. If $\mathbbm{L}\cong \mathbbm{R}$ the variable is said \emph{continuous}.
\end{defn}

\begin{justification}
	We claim $\mathbbm{L}$ is a topological space. $\mathbbm{L}$ is a set of physically distinguishable possibilities. $\mathbbm{L}$ is a topological space because of \ref{topologically_distinguishable}.

	We claim $q$ is a continuous map. $q$ is a map between two sets of physically distinguishable elements. $q$ is continuous because of \ref{continuous_map}.
\end{justification}

When combining multiple state variables, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen item, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have two values. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=s$ is available; for $n=2$ we can choose $l=s$ or $l=p$. The choices are not independent. That is: two or more variables are independent if there always exists a state for any possible combination, if the total number of states is the cartesian product of the possibilities of each variable. We call \emph{state vector} a collection of independent state variables that fully identify a state.

@@

@@

The discrete variable do not present such problems, as any number of them can be flattened out in a single one (i.e. $\mathbbm{Z}^n$ is homeomorphic to $\mathbbm{Z}$). They will determine the number of connected components of the manifold. Once we introduce a continuous parameter for time evolution, though, these also become irrelevant as continuous time evolution requires continuous trajectories that cannot move states across disconnected components. This justifies the special interest in path connected manifolds, as this is where trajectories for deterministic and reversible continuous evolution live.

@@

\begin{proof}
	We claim $\mathcal{S}'$ is a manifold. Let $\mathcal{s} \in \mathcal{S}$. Let $n$ be the number of independent continuous variables needed to identify $\mathcal{s}$. There exists a neighborhood $U$ around $\mathcal{s}$ where we have $(q_1,...,q_n):U\rightarrow \mathbbm{R}^n$. This map is a bijection as $s$ is identified by those variables. $\mathcal{S}$ is homeomorphic to $\mathbbm{R}^n$ around  $\mathcal{s}$. Let $\mathcal{S}'$ be the set of all states potentially reachable by deterministic and reversible evolution from $\mathcal{s}$. Deterministic and reversible evolution is a homeomorphism between initial and final states. $\forall \hat{\mathcal{s}} \in \mathcal{S}'$ the evolution of $U$ will contain $\hat{\mathcal{s}}$ and will be isomorphic to $\mathbbm{R}^n$. $\mathcal{S}'$ equipped with the subspace topology is a Hausdorff, second countable topological space everywhere homeomorphic to $\mathbbm{R}^n$. $\mathcal{S}'$ is a manifold of dimension $n$.

	We claim $\mathcal{S}'$ is path connected. Let $\hat{\mathcal{s}} \in \mathcal{S}'$. There exist map $\lambda : [t_0,t_1] \rightarrow \mathcal{S}'$ where $\lambda(t_0)=\mathcal{s}$, $\lambda(t_1)=\hat{\mathcal{s}}$ and $t_0$ and $t_1$ are the initial and final time respectively. $\lambda$ is a map from two physically distinguishable quantities, time and states. $\lambda$ is continuous by \ref{continuous_map}. All $\hat{\mathcal{s}} \in \mathcal{S}'$ are path connected to $\mathcal{s}$. $\mathcal{S}'$ is path connected.
\end{proof}

\subsection{Evolution of state variables}
